\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-18 rebuttal.tex

}
\opening{Dear chair and members of the reading committee}
\todo{fix todos in document}

I would like to extend my sincere gratitude for time and effort you have taken
to review the manuscript of my PhD dissertation.
Additionally, I would like to thank you for the insightful, constructive and
detailed comments you all have provided.
The following reply contains a point-by-point response to your review comments.
Attached, you will also find an updated draft of the manuscript and a version
with indicated changes for your approval.

\reply{Andrea K\"ubler}
Major comments:
\begin{enumerate}
	\item
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item
\end{enumerate}

\reply{Fabien Lotte}
Major comments:
\begin{enumerate}
	\item Part of the confusion between active and passive BCI might be due
	to the poor choice of phrasing ``paradigms that require little active
  participation from the user, like redirecting attention or initiating imagined actions, are classified as passive BCI.''.
	This has been rewritten to ``in contrast to paradigms that require the user's active participation, for
	instance through redirecting attention or initiating imagined actions,
	paradigms free of this requirement are classified as passive bci,''
	for clarity.
	Neurofeedback has been reclassified as active BCI in the text and in
	figure 1.3.
	\item \todo{Why XDAWN+RG no shrinkage (performed worse), fair
	comparison of sample rate}
  \item Appendix 1.A now fully reports all obtained accuracies to give a better
    insight in effect sizes. Wherever the significant effects have been mentioned
    in the text, the effect size has now been reported along with it.
	\item \todo{report actual accuracies performance stbf in table}
	\item In the results section of chapter 4, statistical comparisons were
	only performed between methods for which we evaluated the performance
	ourselves (HODA, PARAFACDA and BTTDA).
	Since the MOABB benchmarking framework~\cite{Aristimunha2023} allows for
	fair comparison of performance metrics reported across different works,
	the analyses for other comparison methods were not re-run
	by us as reported, but performances were
	taken from the MOABB benchmark database~\cite{Chevallier2024}.
	However, this source provided only the aggregated performance metrics
	and not the metrics per subject, session and cross-validation fold,
	hence we could not perform relevant statistical comparison with these
	comparison models.
	\item \todo{include table with absolute performances for transfer
	conditions}
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item According to the annotations, typographical errors have been
	corrected in text and figures, and imprecise or inaccurate statements have been adjusted.
	The recommended references have been discussed and, where necessary,
	statements were updated accordingly.
  Figure 1.5 was corrected and matched with the correct caption.
\end{enumerate}

### 2024-11-18 chapters/bci.tex
Add accuracies table

Specifically, invasive methods are less noisy and come with an increased \emph{spatial
resolution}, meaning they can extract a signal from a specific brain region or even a
set of neurons of interest.
This makes it easier to retrieve a signal with a known origin from the brain, which
might be leveraged by the paradigm to modulate information.
Generally, microelectrodes penetrating the cortex are considered to have the highest
spatial resolution.
These are needles with a diameter smaller than 10 µm and one or multiple electrodes that
extend several millimeters into the cortex.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular potential
of a small group of neurons,\todo{qualitative}
or even intracellular single neuron action potential
spikes.
They come in single-electrode form, or, more popularly, in microelectrode arrays.
Well-known examples are the Utah array~\cite{Maynard1997}, the more recent NeuraLink
implant~\cite{Musk2019}, and IMEC's Neuropixels 2.0~\cite{Steinmetz2021}.

Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
Depth electrodes, like those found in in stereo-\ac{eeg}, can be used for \ac{bci}~\cite{Wu2024}
 and other applications, like closed-loop adaptive control for Deep Brain Stimulation to mitigate Parkinson's disease
symptoms~\cite{Arlotti2018}.\todo{qualify size}
Depth electrodes are longer rods with multiple electrodes that can extend deeper into
the brain.
Since these still penetrate the cortex, electrodes that sit on top of the cortex are
more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and
epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG), with a higher
number of smaller electrodes per surface area (recording from sites with
diameter of 10s of micrometers instead of $\pm$ 1 cm)~\cite{Shokoueinejad2019},
allow for more precise measurements.

>>>>

Specifically, invasive methods are less noisy and come with an increased \emph{spatial
resolution}, meaning they can extract a signal from a specific brain region or even a
set of neurons of interest.
This makes it easier to retrieve a signal with a known origin from the brain, which
might be leveraged by the paradigm to modulate information.
Generally, microelectrodes penetrating the cortex are considered to have the highest
spatial resolution.
These are needles with a diameter smaller than 10 µm and one or multiple electrodes that
extend several millimeters into the cortex.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular potential
of a small group of neurons, or even intracellular single neuron action potential
spikes.
They come in single-electrode form, or, more popularly, in microelectrode arrays.
Well-known examples are the Utah array~\cite{Maynard1997}, the more recent NeuraLink
implant~\cite{Musk2019}, and IMEC's Neuropixels 2.0~\cite{Steinmetz2021}.

Larger implanted electrodes (usually a few millimeters in size) are referred to
as intracranial \ac{eeg} (i\ac{eeg}).
Depth electrodes, like those found in in stereo-\ac{eeg}, can be used for \ac{bci}~\cite{Wu2024}
 and other applications, like closed-loop adaptive control for Deep Brain Stimulation
 to mitigate Parkinson's disease
symptoms~\cite{Arlotti2018}.
Depth electrodes are longer rods with multiple electrodes that can extend deeper into
the brain.
Since these still penetrate the cortex, electrodes that sit on top of the cortex are
more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and
epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG), with a higher
number of smaller electrodes per surface area (recording from sites with
diameter of 10s of micrometers instead of $\pm$ 1 cm)~\cite{Shokoueinejad2019},
allow for more precise measurements.

### 2024-11-18 chapters/gaze_independence.tex
Add accuracies table

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but they often do so only when paired
with invasive recording.
Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010}
stimulation do not rely on gaze redirection but also result in lower \acp{itr}
compared to visual paradigms\todo{qualitative}.
While auditory and somatosensory stimulation on their own might yield poor
\ac{itr}, \textcite{Yin2016} showed that it can provide added value in gaze-independent
settings when coupled with oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
In general, their approach is interesting as it follows a philosophy that
relies on establishing as many communication pathways as possible.
Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
operation and from user-dependent
variability~\cite{Severens2014,Reichert2020b}.\todo{qualitative}
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

@@

More recent work has made advances in decoding lateralized covert
\ac{vsa} by harnessing the N2pc \ac{erp}
component~\cite{Thiery2016,Reichert2020b,Wang2022}.
The amplitude of the N2pc component in the contralateral hemisphere is directly modulated
by the location of covert \ac{vsa}.
Another approach is directly decoding covert shifts in \ac{vsa} from spectral
content~\cite{Tonin2013}.
These methods are promising, but to date often require a slower stimulation pace or cannot display as
many targets as the classical \ac{erp} \ac{bci} paradigms.
As with the alternative modality approach, they are currently only helpful as
extra communication channels in a hybrid paradigm.
\textcite{Xu2016} reached an \ac{itr} of 23.56 bits/min by detecting the N2pc
component while performing \ac{ssvep} stimulation.
\textcite{Egan2017} also showed an increase in \ac{ssvep} performance by
concurrently decoding the covert \ac{vsa} shift from spectral content, but
their study only used a few stimulation targets.\todo{qualitative}
In general, we can state that gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
leveraging spatial attention remains underexplored.

@@

Due to the decreased amplitude of the N1 and P3 components in covert attention
settings~\cite{Treder2010}, the \ac{snr} of
the \ac{erp} is lower than in overt attention settings\todo{qualitative}.
Therefore, a straightforward way to reach satisfactory
gaze-independent decoding performance may be to increase overall \ac{erp}
decoding performance.

>>>>

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but they often do so only when paired
with invasive recording.
Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010}
stimulation do not rely on gaze redirection but also result in lower \acp{itr}
compared to visual paradigms, with \acp{itr} ranging between 1 and 10 bits/min
for auditory paradigms and 0.5 and 5 bits/min for tactile~\cite{Riccio2012}.
While auditory and somatosensory stimulation on their own might yield poor
\ac{itr}, \textcite{Yin2016} showed that it can provide added value in gaze-independent
settings when coupled with oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
In general, their approach is interesting as it follows a philosophy that
relies on establishing as many communication pathways as possible.
Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
operation and from user-dependent
variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

@@

More recent work has made advances in decoding lateralized covert
\ac{vsa} by harnessing the N2pc \ac{erp}
component~\cite{Thiery2016,Reichert2020b,Wang2022}.
The amplitude of the N2pc component in the contralateral hemisphere is directly modulated
by the location of covert \ac{vsa}.
Another approach is directly decoding covert shifts in \ac{vsa} from spectral
content~\cite{Tonin2013}.
These methods are promising, but to date often require a slower stimulation pace or cannot display as
many targets as the classical \ac{erp} \ac{bci} paradigms.
As with the alternative modality approach, they are currently only helpful as
extra communication channels in a hybrid paradigm.
\textcite{Xu2016} reached an \ac{itr} of 23.56 bits/min by detecting the N2pc
component while performing \ac{ssvep} stimulation.
\textcite{Egan2017} also showed an increase in \ac{ssvep} performance by
concurrently decoding the covert \ac{vsa} shift from spectral content, but
their study only used two stimulation targets.
In general, we conclude that gaze-independent decoding in the fast-paced reactive
\ac{erp} paradigm leveraging spatial attention remains underexplored.

@@

Due to the decreased amplitude of the N1 and P3 components in covert attention
settings~\cite{Treder2010}, the \ac{snr} of
the \ac{erp} is lower than in overt attention settings.
Therefore, a straightforward way to reach satisfactory
gaze-independent decoding performance may be to increase overall \ac{erp}
decoding performance.

### 2024-11-18 chapters/stbf_struct.tex
Add accuracies table

	The \ac{eeg} recorded at 2048Hz was re-referenced off-line to the average of the mastoids.
	The reference electrodes were dropped from the analysis.
	Data were subsequently filtered between 0.5Hz and 16Hz using forward-backward
	filtering with a fourth-order Butterworth IIR filter.
	The \ac{eeg} signal was corrected for ocular artifacts using Independent Component
	Analysis (ICA) by rejecting components that correlated with the bipolar EOG
  channels vEOG and hEOG, according to adaptive Z-score
  thresholding.\todo{details? threshold for rejection?}
	Finally, epochs were cut from 0.1s to 0.6s after stimulus onset.
	No baseline correction was performed since this affects the temporal covariance
	of the data, violating the Toeplitz structure assumption~\cite{Bijma2003}.

@@

	\cref{fig:iterations} confirms these results for the \textsc{stbf-struct} classifier.
	Using more than one fixed-point iteration does not significantly improve the
	accuracy across the amounts of training data and the number of trials
  used for evaluation\todo{qualitative}.
	Hence, only one iteration is used for the \textsc{stbf-struct} classifier, leading to a drastic speed-up of
  the training process\todo{qualitative}.

@@

  \begin{table}[p]
		\centering
    \makebox[\textwidth][c]{%
		\input{chapters/stbf_struct_tests_1.tex}
  }
    \caption[Statistical significance of differences in classifier performance (1 repetition).]{%
      Statistical significance of differences in classifier performance.
      $p$-values calculated by one-sided paired Wilcoxon rank-sum test with
			Holm correction using one testing trial for different classifiers
			and levels of data availability. $p$-values $<0.05$ are considered significant
			and marked in grey.}
		\label{tab:stbf-struct/p-values-1}
    \bigskip

		\centering
    \makebox[\textwidth][c]{%
		\input{chapters/stbf_struct_tests_2.tex}
    }
		\caption[Statistical significance of differences in classifier performance (2 repetitions).]{%
      $p$-values as in \cref{tab:stbf-struct/p-values-1} averaging over two testing trials.}
		\label{tab:stbf-struct/p-values-2}
    \bigskip

		\centering
    \makebox[\textwidth][c]{%
		\input{chapters/stbf_struct_tests_3.tex}
    }
    \caption[Statistical significance of differences in classifier performance (5 repetitions).]{%
      $p$-values as in \cref{tab:stbf-struct/p-values-1} averaging over five testing trials.}
		\label{tab:stbf-struct/p-values-5}
	\end{table}
	\unskip

@@

	\begin{figure}
		%\includegraphics[width=.66\linewidth]{figures/stbf_struct/training_time.eps}
    \sffamily
    \sansmath
    \input{figures/stbf_struct/training_time.pgf}
    \caption[Classifier training time.]{Training time boxplots of different
    classifiers and individual results for all training folds for all subjects
    using 32 channels. Using structured covariance estimation greatly reduces
    training time.}
		\label{fig:stbf-struct/training-time}
	\end{figure}
  \todo{the legend does not correspond to the text (method names are
  different)}

	\cref{fig:stbf-struct/training-time} shows that the training time of
	\textsc{stbf-struct} has a lower median training time than
  \textsc{stbf-shrunk}, \textsc{stbf-emp} and \textsc{xdawn+rg} when training
  on all available data and using 32 channels\todo{qualitative}.

@@

>>>>

	The \ac{eeg} recorded at 2048Hz was re-referenced off-line to the average of the mastoids.
	The reference electrodes were dropped from the analysis.
	Data were subsequently filtered between 0.5Hz and 16Hz using forward-backward
	filtering with a fourth-order Butterworth IIR filter.
	The \ac{eeg} signal was corrected for ocular artifacts using Independent Component
	Analysis (ICA) by rejecting components that correlated with the bipolar EOG
  channels vEOG and hEOG using iterative Z-score thresholding.
  Components with a Z-scored pearson correlation coefficient
  exceeding the other 3 times the standard deviation of the Z-scored
  correlation coefficients of other component's Z-scored were iteratively
  rejected until none that exceed the threshold remained.
  Finally, epochs were cut from 0.1s to 0.6s after stimulus onset.
	No baseline correction was performed since this affects the temporal covariance
	of the data, violating the Toeplitz structure assumption~\cite{Bijma2003}.

@@

	\cref{fig:iterations} confirms these results for the \textsc{stbf-struct} classifier.
	Using more than one fixed-point iteration does not significantly improve the
	accuracy across the amounts of training data and the number of trials
  used for evaluation\todo{qualitative}.
	Hence, only one iteration is used for the \textsc{stbf-struct} classifier, leading to a drastic speed-up of
  the  training process.

@@

@@

	\begin{figure}
		%\includegraphics[width=.66\linewidth]{figures/stbf_struct/training_time.eps}
    \sffamily
    \sansmath
    \input{figures/stbf_struct/training_time.pgf}
    \caption[Classifier training time.]{Training time boxplots of different
    classifiers and individual results for all training folds for all subjects
    using 32 channels. Using structured covariance estimation greatly reduces
    training time.}
		\label{fig:stbf-struct/training-time}
	\end{figure}

	\Cref{fig:stbf-struct/training-time} shows that the training time of
	\textsc{stbf-struct} has a lower median training time than
  \textsc{stbf-shrunk}, \textsc{stbf-emp} and \textsc{xdawn+rg} when training
  on all available data and using 32 channels\todo{qualitative}.

@@

\clearpage
\begin{subappendices}
  \section{Accuracies per number of trials and available training data}

  \footnotesize
  \begin{table}[!h]
	    \input{chapters/stbf_struct-acc-stbf_kron_toep_shrunk.tex}
      \caption{Accuracies for \textsc{stbf-struct}.}
  \end{table}
  \begin{table}[!h]
	    \input{chapters/stbf_struct-acc-stbf_loocv.tex}
      \caption{Accuracies for \textsc{stbf-shrunk}.}
  \end{table}
  \begin{table}[!h]
	    \input{chapters/stbf_struct-acc-stbf_emp.tex}
      \caption{Accuracies for \textsc{stbf-emp}.}
  \end{table}
  \begin{table}[!h]
	    \input{chapters/stbf_struct-acc-xdawn_rg.tex}
      \caption{Accuracies for \textsc{xdawn+rg}.}
  \end{table}

  \clearpage

  \clearpage
  \section{Statistical comparisons}
  \begin{table}[h]
    \footnotesize
    \begin{subtable}{\linewidth}
	    \input{chapters/stbf_struct_tests_1.tex}
	    \caption{Evaluation on one testing trial.}
      \label{tab:stbf-struct/p-values-1}
    \end{subtable}
    \bigskip

    \begin{subtable}{\linewidth}
	    \input{chapters/stbf_struct_tests_2.tex}
	    \caption{Evaluation averaging over two testing trials.}
	    \label{tab:stbf-struct/p-values-2}
    \end{subtable}
    \bigskip

    \begin{subtable}{\linewidth}
	    \input{chapters/stbf_struct_tests_3.tex}
	    \caption{Evaluation averaging over five testing trials.}
	    \label{tab:stbf-struct/p-values-5}
    \end{subtable}
    \caption[Statistical significance of differences in classifier
    performance.]{%
        Statistical significance of differences in classifier performance.
        $p$-values calculated by one-sided paired Wilcoxon rank-sum test with
	    	Holm correction using one testing trial for different classifiers
	    	and levels of data availability. $p$-values $<0.05$ are considered
        significant.
    }
	\end{table}

\end{subappendices}
